Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex

A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.

The most widely used \<Map> implementations have the following properties:
\begin{itemize}
    \item \<HashMap> is implemented in terms of a hash table, which never
      guarantees deterministic iteration over its entries. A \codeid{@Det
        HashMap} does not exist.
    \item \<LinkedHashMap>, like \<List>, can have any of the \<@NonDet>,
      \<@OrderNonDet>, or \<@Det> type qualifiers. Iterating over a
      \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet
    LinkedHashMap> can be created by calling its constructor with a
    \<@OrderNonDet HashMap>.
    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
      TreeMap> doesn't exist because the entries are always ordered.
\end{itemize}

@@

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
the type of \<identity(arg)> is the same as the type of
\<arg>.  A programmer can also write a type qualifier on a type argument,
as in \codeid{@NonDet List<@NonDet Integer>}.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
interface ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
for ``polymorphic determinism qualifier.)
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

>>>>

A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
\Cref{code-determinism} shows that the determinism qualifier on the return type of \<entrySet()> is the
same as that on the receiver.  That is, its type (ignoring type arguments) is
$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .

We have not yet found any implementation of the \<Map>
interface in which \<entrySet()> \<keySet()>, and \<values()> behave
differently with respect to determinacy:  either all, or none, or deterministic.

The most widely used \<Map> implementations have the following properties:
\begin{itemize}
    \item \<HashMap> is implemented in terms of a hash table, which never
      guarantees deterministic iteration over its entries. A \codeid{@Det
        HashMap} does not exist.
    \item \<LinkedHashMap>, like \<List>, can have any of the \aNonDet,
      \aOrderNonDet, or \aDet type qualifiers. Iterating over a
      \<LinkedHashMap> returns
    its entries in the order of their insertion. An \aOrderNonDet\
    \<LinkedHashMap> can be created by passing an
    \aOrderNonDet\ \<HashMap> to its constructor, as in \<new
    LinkedHashMap(myOndHashMap)>.
    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
      TreeMap> doesn't exist because the entries are always sorted.
\end{itemize}

@@

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
the type of \<identity(arg)> is the same as the type of
\<arg>.  A programmer can also write a type qualifier on a type argument,
as in \codeid{@NonDet List<@NonDet Integer>}.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
interface ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
for ``polymorphic determinism qualifier.)
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but \emph{not} on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}
